---
title: "Analysis of computational simulations of control data in the 80S-hit model"
author: "`r Sys.info()[['user']]`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  github_document:
    toc: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

# Load libraries
```{r}
library(tidyverse)
library(rasilabRtemplates)
library(viridis)
library(dplyr)
# disable scientific notation
options(scipen=999)
```

# Read reaction firing data
```{r}
rxn_data <- list.files("output/", pattern = "rxn_list.tsv.gz", full.names = T) %>% 
  enframe() %>% 
  mutate(sim_id = as.integer(str_extract(value, "(?<=model_)\\d+"))) %>% 
  mutate(data = map(value, read_tsv)) %>%
  select(-name, -value) %>% 
  unnest() %>%
  print()
```

# Read simulation time data
```{r}
sim_data <- list.files("output/", pattern = "molecule_type_list.tsv.gz", full.names = T) %>% 
  enframe() %>% 
  mutate(sim_id = as.integer(str_extract(value, "(?<=model_)\\d+"))) %>% 
  mutate(data = map(value, read_tsv)) %>%
  select(-name, -value) %>% 
  unnest() %>% 
  filter(mol_type %in% c("simulated_time", "cpu_time")) %>% 
  spread(mol_type, mol_type_id) %>% 
  print()
```

# Read last reaction time data

```{r}
last_rxn_data <- list.files("output/", pattern = "molecule_type_list.tsv.gz", full.names = T) %>% 
  enframe() %>% 
  mutate(sim_id = as.integer(str_extract(value, "(?<=model_)\\d+"))) %>% 
  mutate(data = map(value, read_tsv)) %>%
  select(-name, -value) %>% 
  unnest() %>% 
  filter(mol_type == "last_rxn_firing_time") %>%
  spread(mol_type, mol_type_id) %>% 
  print()
```

# Check to make sure last reaction fired near end of simulation. 
0 rows output means all simulations' last reactions fired in last 0.1% of simulated time

```{r}
check_rxn <- last_rxn_data %>%
  left_join(sim_data, by = "sim_id") %>%
  mutate(ratio = last_rxn_firing_time / simulated_time) %>%
  filter(ratio < 0.999) %>%
  print()
```

# Read simulation parameters
```{r}
annotations  <- read_tsv('sim.params.tsv') %>%
  mutate(uorf2_length = (uorf2_stop - uorf2_start)/3) %>%
  print()
```

# Calculate protein synthesis rate of main ORF

```{r}
psr_data <- rxn_data %>% 
  mutate(mrna_loc = str_extract(name, "\\d+$")) %>% 
  left_join(annotations %>% select(sim_id, uorf2_stop, orf_stop), by = "sim_id") %>% 
  filter((mrna_loc == orf_stop) & str_detect(name, "terminate")) %>% 
  group_by(sim_id, mrna_loc) %>% 
  dplyr::summarize(n_firings = sum(n_firings)) %>% 
  left_join(sim_data, by = "sim_id") %>% 
  mutate(psr = n_firings / simulated_time) %>% 
  select(sim_id, mrna_loc, psr) %>% 
  print()
```

# Get only annotations that we want to use for plotting

```{r}
param_data <- annotations %>% 
    select(sim_id, k_cap_bind, k_elong_stall, k_start_uorf2, k_scan_term_5_hit_80s, k_terminated_ssu_recycle_uorf2, uorf2_length) %>%
  print()
```

# Combine all processed data into a single table
```{r}
data <- param_data %>% 
  left_join(psr_data, by = "sim_id") %>%
  dplyr::rename(k_recycle = k_terminated_ssu_recycle_uorf2,
         k_abort = k_scan_term_5_hit_80s
         ) %>% 
  mutate(k_recycle = round(k_recycle, 1)) %>%
  print()
```

# Plot PSR as a function of initiation rate for different rates of uORF2 (re)initiation, length
```{r,fig.width=8,fig.height=3}
plot_data <- data %>%
  print()

plot_data %>% 
  ggplot(aes(x = log2(k_cap_bind), y = log2(psr), color = as.factor(k_elong_stall),  group = k_elong_stall)) +
  facet_wrap(~ k_start_uorf2 + k_recycle + uorf2_length, ncol = 3, labeller = label_both, scales = "free") +
  geom_point() +
  geom_line() +
  scale_color_manual(values = cbPalette[c(2,1)])

```

# Read in control mutants experimental data for fitting

Based on standard error, if we fit within log2(0.1), that will be good enough.
```{r}
expt_data <- read_csv("tables/nluc_fluc_control_mutants.csv") %>% 
  mutate(label = c("wt", "no_stall", "no_aug", "strong")) %>% 
  select(label, mean_ratio) %>%
  deframe() %>%
  print()
```

# Perform a spline fit of PSR vs k_cap_bind for interpolation and fitting to experimental data

We fit a spline and evaluate it at 100 points so that we can get pretty close to the experimental value

```{r}
fit_data <- data %>% 
  group_by(across(starts_with("k_")), uorf2_length) %>% 
  ungroup(k_cap_bind) %>% 
  nest() %>% 
  mutate(fit = map(data, function(df) spline(x = df$k_cap_bind, y = df$psr, n = 100) %>% as.data.frame)) %>% 
  select(-data) %>%
  unnest(fit) %>%
  rename(k_cap_bind = x, psr = y) %>%
  print()
```

# Extract the PSR of WT, strong Kozak, no stall w.r.t to no aug in the simulation and minimize the RMS error
```{r}
fit_data %>% 
  ungroup() %>%
  # convert stall to be a binary factor
  mutate(stall = if_else(k_elong_stall == 0.001, T, F)) %>% 
  # convert uORF2 start strength to be binary
  mutate(kozak = if_else(k_start_uorf2 == 0.1, "WT", if_else(k_start_uorf2 == 0, "no_aug", "strong"))) %>% 
  # assign mutant to the same names as experiments for easy comparison
  mutate(mutant = case_when(
    stall == T & kozak == "WT" ~ 'wt',
    stall == T & kozak == "strong" ~ 'strong',
    stall == F & kozak == "WT" ~ 'no_stall',
    stall == T & kozak == "no_aug" ~ "no_aug")) %>% 
  # get rid of the of mutants that we are not minimizing (eg. no stall + no aug/strong kozak)
  filter(!is.na(mutant)) %>% 
  select(-stall, -kozak, -k_elong_stall, -k_start_uorf2) %>%
  pivot_wider(names_from = "mutant", values_from = "psr") %>%
  # calculate the log2 difference between no_stall and wt / strong
  mutate(wt = log2(wt) - log2(no_aug)) %>% 
  mutate(strong = log2(strong) - log2(no_aug)) %>%
  mutate(no_stall = log2(no_stall) - log2(no_aug)) %>%
  # calculate RMS error
  mutate(rms = (wt - expt_data['wt'])^2 + (strong - expt_data['strong'])^2) %>% 
  # minimize RMS error in each model and for other parameters as needed
  arrange(rms) %>%
  slice(1) %>%
  write_csv("./tables/fit_params.csv") %>%
  knitr::kable()

```
