---
title: "Analysis of computational simulations of the regulated re-initiation model"
author: "`r Sys.info()[['user']]`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  github_document:
    toc: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

# Load libraries
```{r}
library(tidyverse)
library(rasilabRtemplates)
library(viridis)
# disable scientific notation
options(scipen=999)
```

# Read reaction firing data
```{r}
rxn_data <- list.files("output/", pattern = "rxn_list.tsv.gz", full.names = T) %>% 
  enframe() %>% 
  mutate(sim_id = as.integer(str_extract(value, "(?<=model_)\\d+"))) %>% 
  mutate(data = map(value, read_tsv)) %>%
  select(-name, -value) %>% 
  unnest() %>%
  print()
```

# Read simulation time data
```{r}
sim_data <- list.files("output/", pattern = "molecule_type_list.tsv.gz", full.names = T) %>% 
  enframe() %>% 
  mutate(sim_id = as.integer(str_extract(value, "(?<=model_)\\d+"))) %>% 
  mutate(data = map(value, read_tsv)) %>%
  select(-name, -value) %>% 
  unnest() %>% 
  filter(mol_type %in% c("simulated_time", "cpu_time")) %>% 
  spread(mol_type, mol_type_id) %>% 
  print()
```

# Read last reaction time data

```{r}
last_rxn_data <- list.files("output/", pattern = "molecule_type_list.tsv.gz", full.names = T) %>% 
  enframe() %>% 
  mutate(sim_id = as.integer(str_extract(value, "(?<=model_)\\d+"))) %>% 
  mutate(data = map(value, read_tsv)) %>%
  select(-name, -value) %>% 
  unnest() %>% 
  filter(mol_type == "last_rxn_firing_time") %>%
  spread(mol_type, mol_type_id) %>% 
  print()
```

# Check to make sure last reaction fired near end of simulation. 
0 rows output means all simulations' last reactions fired in last 0.1% of simulated time

```{r}
check_rxn <- last_rxn_data %>%
  left_join(sim_data, by = "sim_id") %>%
  mutate(ratio = last_rxn_firing_time / simulated_time) %>%
  filter(ratio < 0.999) %>%
  print()
```

# Read simulation parameters
```{r}
annotations  <- read_tsv('sim.params.tsv') %>%
  print()
```

# Calculate protein synthesis rate of main ORF

```{r}
psr_data <- rxn_data %>% 
  mutate(mrna_loc = str_extract(name, "\\d+$")) %>% 
  left_join(annotations %>% select(sim_id, uorf2_stop, orf_stop), by = "sim_id") %>% 
  filter((mrna_loc == orf_stop) & str_detect(name, "terminate")) %>% 
  group_by(sim_id, mrna_loc) %>% 
  dplyr::summarize(n_firings = sum(n_firings)) %>% 
  left_join(sim_data, by = "sim_id") %>% 
  mutate(psr = n_firings / simulated_time) %>% 
  select(sim_id, mrna_loc, psr) %>% 
  print()
```

# Get only annotations that we want to use for plotting

```{r}
param_data <- annotations %>%
  mutate(uorf2_reinit = round(100 * 5 / (k_terminated_ssu_recycle_uorf2 + 5)),
         uorf3_reinit = round(100 * 5 / (k_terminated_ssu_recycle_uorf3 + 5)),
         uorf2_init = round_any(100 * k_start_uorf2 / (k_start_uorf2 + 5), 10),
         uorf3_init = round_any(100 * k_start_uorf3 / (k_start_uorf3 + 5), 10)) %>%
  select(sim_id, uorf2_init, uorf3_init, uorf2_reinit, uorf3_reinit, n_tc_0) %>%
  print()
```

# Combine all processed data into a single table
```{r}
data <- param_data %>% 
  left_join(psr_data, by = "sim_id") %>%
  print()
```

# Plot PSR as a function of uorf2/3 (re)init %s --> buffering requires high uORF2 reinit, low uORF3 reinit, high uORF3 init, ~agnostic to uORF2 init (as long as its not super low)

```{r,fig.width=3,fig.height=2.25}
plot_data <- data %>%
  filter(uorf2_init %in% c(0, 100),
         uorf3_init %in% c(0, 100),
         uorf2_reinit != uorf3_reinit,
         uorf2_reinit %in% c(1, 100),
         uorf3_reinit %in% c(1, 100),
         uorf2_init != 0 | uorf3_init !=0) %>%
  print()

reinit2_labeller <- function(x) paste0("uORF1\ncontinued\nscanning: ", x, "%")
reinit3_labeller <- function(x) paste0("uORF2\ncontinued\nscanning: ", x, "%")

plot_data %>% 
  ggplot(aes(x = log2(n_tc_0), y = log2(psr), color = interaction(uorf2_init, uorf3_init))) +
  facet_wrap(~ uorf2_reinit + uorf3_reinit, ncol = 3, labeller = labeller(uorf2_init = init2_labeller, uorf2_reinit = reinit2_labeller, uorf3_reinit = reinit3_labeller)) +
  geom_point(size = 1) +
  geom_line() +
  scale_color_viridis(discrete = T, end = 0.9) +
  scale_y_continuous(breaks = scales::pretty_breaks(n=3)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n=3)) +
  theme(axis.text.y = element_text(size = 7), axis.text.x = element_text(size = 7)) +
  labs(y = "main ORF protein\noutput (log2, s-1)", x = "ternary complex concentration (log2)", color = "Baseline\nuORF1.\nuORF2\ninitiations (%)")

  ggsave('figures/hinnebusch_buffering_main.pdf')
```

# Get reactions at start codons

```{r}

start_codon_rxns <- rxn_data %>%
  filter(str_detect(name, "^(scan_[:digit:]|scan_from_|scan_to)")) %>%
  # uORF2/3 and main ORF start at 25, 97, 232
  filter(str_detect(name, c("_25$|_97$|_232$"))) %>%
  select(sim_id, name) %>%
  print()

```

# Get reactions at stop codons

```{r}

stop_codon_rxns <- rxn_data %>%
  filter(str_detect(name, "^(scan_[:digit:]|scan_from_|scan_to|recycle_[:digit:]|terminate_[:digit:]|recycle_from)")) %>%
  # uORF2/3 stop at 91, 106
  filter(str_detect(name, c("_91$|_106$"))) %>%
  select(sim_id, name) %>%
  print()

```

# Get the number of scanning and initiating ribosomes at start codons join with parameters

```{r}
start_codon_data_raw <- rxn_data %>% 
  right_join(start_codon_rxns, by = c("sim_id", "name")) %>%
  select(sim_id, n_firings, name) %>% 
  spread(name, n_firings) %>% 
  left_join(param_data, by = c("sim_id")) %>%
  print()
  # uORF2/3 and main ORF start at 25, 97, 232
```

# Get the number of scanning and initiating ribosomes at stop codons join with parameters

```{r}
stop_codon_data_raw <- rxn_data %>% 
  right_join(stop_codon_rxns, by = c("sim_id", "name")) %>%
  select(sim_id, n_firings, name) %>% 
  spread(name, n_firings) %>% 
  left_join(param_data, by = c("sim_id")) %>%
  print()
  # uORF2/3 stop at 91, 106
```

# Get the fraction of scanning and initiating ribosomes at start codons

```{r,fig.width=2.25,fig.height=1.25}
start_codon_data <- start_codon_data_raw %>%
  # uORF2/3 and main ORF start at 25, 97, 232
  mutate(total_25 = scan_25 + scan_to_elongate_25 + scan_to_elongate_from_leading_collision_25) %>%
  mutate(total_97 = scan_97 + scan_from_elongation_collision_97 + scan_from_scan_collision_97 + scan_to_elongate_97 + scan_to_elongate_from_leading_collision_97 + scan_to_elongate_from_trailing_elongation_collision_97 + scan_to_elongate_from_trailing_scan_collision_97) %>%
  mutate(total_232 = scan_232 + scan_from_elongation_collision_232 + scan_from_scan_collision_232 + scan_to_elongate_232 + scan_to_elongate_from_leading_collision_232 + scan_to_elongate_from_trailing_elongation_collision_232 + scan_to_elongate_from_trailing_scan_collision_232) %>%
  mutate(initiate_25 = round((100*(scan_to_elongate_25 + scan_to_elongate_from_leading_collision_25) / total_25), 1)) %>%
  mutate(initiate_97 = round((100*(scan_to_elongate_97 + scan_to_elongate_from_leading_collision_97 + scan_to_elongate_from_trailing_elongation_collision_97 + scan_to_elongate_from_trailing_scan_collision_97) / total_97), 1)) %>%
  mutate(initiate_232 = round((100*(scan_to_elongate_232 + scan_to_elongate_from_leading_collision_232 + scan_to_elongate_from_trailing_elongation_collision_232 + scan_to_elongate_from_trailing_scan_collision_232) / total_232), 1)) %>%
  select(sim_id, uorf2_init, uorf3_init, uorf2_reinit, uorf3_reinit, n_tc_0, initiate_25, initiate_97, initiate_232) %>%
  filter(uorf2_init == 100,
         uorf3_init %in% c(20, 100),
         uorf2_reinit == 100,
         uorf3_reinit == 100) %>%
  print()

start_codon_data %>% 
  ggplot(aes(x = log2(n_tc_0), y = initiate_97, color = as.factor(uorf3_init))) +
  geom_point(size = 1) +
  geom_line() +
  scale_color_viridis(discrete = T, end = 0.9) +
  scale_y_continuous(breaks = scales::pretty_breaks(n=3)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n=3)) +
  theme(axis.text.y = element_text(size = 7), axis.text.x = element_text(size = 7)) +
  labs(y = "uORF2 initiation\n(real, %)", x = "ternary complex\nconcentration (log2)", color = "uORF2\ninitiation\n(predicted, %)")

ggsave('figures/hinnebusch_supp_tc_dependent_init.pdf')

```

# Conclusions

 - preferred translation in the regulated re-initiation model requires several conditions
 1) 2 well-translated uORFs
 2) frequent continued scanning after termination at uORF1
 3) rare continued scanning after termination at uORF2
 - ternary complex concentration-dependent changes in uORF2 initiation underlie this preferred translation effect