---
title: "Analysis of computational simulations of the constitutive repression, queueing-mediated enhanced repression, and collision-mediated 40S dissociation models"
author: "`r Sys.info()[['user']]`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  github_document:
    toc: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

# Load libraries

```{r}
library(tidyverse)
library(rasilabRtemplates)
library(viridis)
library(plyr)
library(dplyr)
# disable scientific notation
options(scipen=999)
```

# Read reaction firing data

```{r}
rxn_data <- list.files("output/", pattern = "rxn_list.tsv.gz", full.names = T) %>% 
  enframe() %>% 
  mutate(sim_id = as.integer(str_extract(value, "(?<=model_)\\d+"))) %>% 
  mutate(data = map(value, read_tsv)) %>%
  select(-name, -value) %>% 
  unnest() %>%
  print()
```

# Read simulation time data

```{r}
sim_data <- list.files("output/", pattern = "molecule_type_list.tsv.gz", full.names = T) %>% 
  enframe() %>% 
  mutate(sim_id = as.integer(str_extract(value, "(?<=model_)\\d+"))) %>% 
  mutate(data = map(value, read_tsv)) %>%
  select(-name, -value) %>% 
  unnest() %>% 
  filter(mol_type %in% c("simulated_time", "cpu_time")) %>% 
  spread(mol_type, mol_type_id) %>% 
  print()
```

# Read last reaction time data

```{r}
last_rxn_data <- list.files("output/", pattern = "molecule_type_list.tsv.gz", full.names = T) %>% 
  enframe() %>% 
  mutate(sim_id = as.integer(str_extract(value, "(?<=model_)\\d+"))) %>% 
  mutate(data = map(value, read_tsv)) %>%
  select(-name, -value) %>% 
  unnest() %>% 
  filter(mol_type == "last_rxn_firing_time") %>%
  spread(mol_type, mol_type_id) %>% 
  print()
```

# Check to make sure last reaction fired near end of simulation. 
0 rows output means all simulations' last reactions fired in last 0.1% of simulated time

```{r}
check_rxn <- last_rxn_data %>%
  left_join(sim_data, by = "sim_id") %>%
  mutate(ratio = last_rxn_firing_time / simulated_time) %>%
  filter(ratio < 0.999) %>%
  print()
```

# Read simulation parameters

```{r}
annotations  <- read_tsv('sim.params.tsv') %>%
  print()
```

# Calculate protein synthesis rate of main ORF

```{r}
psr_data <- rxn_data %>% 
  mutate(mrna_loc = str_extract(name, "\\d+$")) %>%
  left_join(annotations %>% select(sim_id, uorf2_stop, orf_stop), by = "sim_id") %>% 
  filter((mrna_loc == orf_stop) & str_detect(name, "terminate")) %>% 
  group_by(sim_id, mrna_loc) %>% 
  dplyr::summarize(n_firings = sum(n_firings)) %>%
  left_join(sim_data, by = "sim_id") %>% 
  mutate(psr = n_firings / simulated_time) %>% 
  select(sim_id, mrna_loc, psr) %>%
  print()
```

# Get only annotations that we want to use for plotting

```{r}
param_data <- annotations %>%
  mutate(uorf_init = round(100*k_start_uorf2 / (5 + k_start_uorf2)),2) %>%
  mutate(uorf_init = round_any(uorf_init, 10)) %>%
  mutate(d_stall = x_stall - uorf2_start) %>%
  mutate(uorf_reinit = 100*5/(k_terminated_ssu_recycle_uorf2 + 5)) %>%
  select(sim_id, k_cap_bind, k_elong_stall, uorf_init, k_scan_term_3_hit_80s, d_stall, uorf_reinit, l_scan_capture) %>%
  print()
```

# Combine all processed data into a single table

```{r}
data <- param_data %>% 
  left_join(psr_data, by = "sim_id") %>%
  dplyr::rename(k_abort = k_scan_term_3_hit_80s) %>% 
  print()
```

# Plot PSR as a function of initiation rate for different rates of uORF2 initiation for constitutive repression model

```{r,fig.width=3,fig.height=1.25}
plot_data <- data %>%
  filter(k_elong_stall == 5,
         k_abort == 0,
         d_stall == 60,
         uorf_reinit %in% c(10, 90),
         l_scan_capture == 0) %>%
  print()

reinit_labeller <- function(x) paste0("re-initiation:\n", x, "%")

plot_data %>% 
  ggplot(aes(x = log2(k_cap_bind), y = log2(psr), color = as.factor(uorf_init))) +
  facet_wrap(~ uorf_reinit, ncol = 2, labeller = labeller(uorf_reinit = reinit_labeller)) +
  geom_point(size = 1) +
  geom_line() +
  viridis::scale_color_viridis(discrete = T, end = 0.9) +
  scale_y_continuous(breaks = scales::pretty_breaks(n=2)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n=3)) +
  theme(axis.text.y = element_text(size = 7), axis.text.x = element_text(size = 7)) +
  labs(x = "ribosome loading (log2, s-1)", y = "main ORF protein\noutput (log2, s-1)",
       color = "Baseline uORF\ninitiation (%)")

ggsave('figures/constitutive_repression_no_buffering.pdf', limitsize = FALSE)

```

# Plot PSR as a function of initiation rate for different rates of uORF2 initiation, d_stall for queuing-mediated enhanced repression model

```{r,fig.width=3,fig.height=1.25}
plot_data <- data %>%
  filter(k_elong_stall == 0.001,
         uorf_init != 0,
         k_abort == 0,
         uorf_reinit == 50,
         l_scan_capture == 0,
         d_stall %in% c(60, 63)) %>%
  print()

d_stall_labeller <- function(x) paste0("d_stall: ", x, " nt")

plot_data %>% 
  ggplot(aes(x = log2(k_cap_bind), y = log2(psr), color = as.factor(uorf_init))) +
  facet_wrap(~ d_stall, ncol = 2, labeller = labeller(d_stall = d_stall_labeller)) +
  geom_point(size = 1) +
  geom_line() +
  viridis::scale_color_viridis(discrete = T, end = 0.9) +
  scale_y_continuous(breaks = scales::pretty_breaks(n=3)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n=3)) +
  theme(axis.text.y = element_text(size = 7), axis.text.x = element_text(size = 7)) +
  labs(x = "ribosome loading (log2, s-1)", y = "main ORF protein\noutput (log2, s-1)",
       color = "Baseline uORF\ninitiation (%)")

ggsave('figures/queuing_mediated_enhanced_repression_buffering_main.pdf', limitsize = FALSE)

```

# Plot PSR as a function of initiation rate for different rates of uORF2 initiation, d_stall for collision-mediated 40S dissociation model

```{r,fig.width=3,fig.height=1.25}
plot_data <- data %>%
  filter(k_elong_stall == 0.001,
         uorf_init != 0,
         k_abort == 2,
         uorf_reinit == 50,
         l_scan_capture == 0,
         d_stall %in% c(60, 63)) %>%
  print()

d_stall_labeller <- function(x) paste0("d_stall: ", x, " nt")

plot_data %>% 
  ggplot(aes(x = log2(k_cap_bind), y = log2(psr), color = as.factor(uorf_init))) +
  facet_wrap(~ d_stall, ncol = 2, labeller = labeller(d_stall = d_stall_labeller)) +
  geom_point(size = 1) +
  geom_line() +
  viridis::scale_color_viridis(discrete = T, end = 0.9) +
  scale_y_continuous(breaks = scales::pretty_breaks(n=3)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n=3)) +
  theme(axis.text.y = element_text(size = 7), axis.text.x = element_text(size = 7)) +
  labs(x = "ribosome loading (log2, s-1)", y = "main ORF protein\noutput (log2, s-1)",
       color = "Baseline uORF\ninitiation (%)")

ggsave('figures/collision_mediated_40S_dissociation_buffering_dstall.pdf', limitsize = FALSE)

```

# Plot PSR as a function of initiation rate for different rates of uORF2 initiation, d_stall for queuing-mediated enhanced repression model

```{r,fig.width=4,fig.height=1.5}
plot_data <- data %>%
  filter(k_elong_stall == 0.001,
         uorf_init != 0,
         k_abort == 0,
         uorf_reinit == 50,
         d_stall %in% c(60, 63),
         d_stall + l_scan_capture != 63) %>%
  print()

d_stall_labeller <- function(x) paste0("d_stall: ", x, " nt")
capture_labeller <- function(x) paste0("l_scan_capture: ", x, " nt")

plot_data %>% 
  ggplot(aes(x = log2(k_cap_bind), y = log2(psr), color = as.factor(uorf_init))) +
  facet_wrap(~ d_stall + l_scan_capture, ncol = 2, scales = "free", labeller = labeller(d_stall = d_stall_labeller, l_scan_capture = capture_labeller)) +
  geom_point(size = 1) +
  geom_line() +
  viridis::scale_color_viridis(discrete = T, end = 0.9) +
  scale_y_continuous(breaks = scales::pretty_breaks(n=2)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n=3)) +
  theme(axis.text.y = element_text(size = 7), axis.text.x = element_text(size = 7)) +
  labs(x = "ribosome loading (log2, s-1)", y = "main ORF protein\noutput (log2, s-1)",
       color = "Baseline uORF\ninitiation (%)")

ggsave('figures/queuing_mediated_enhanced_repression_buffering_supp.pdf', limitsize = FALSE)

```

# Plot PSR as a function of initiation rate for different rates of uORF2 initiation, k_abort for collision-mediated 40S dissociation model

```{r,fig.width=3,fig.height=1.25}
plot_data <- data %>%
  filter(k_elong_stall == 0.001,
         uorf_init != 0,
         d_stall == 63,
         uorf_reinit == 50,
         k_abort %in% c(0, 2),
         l_scan_capture == 0) %>%
  print()

k_abort_labeller <- function(x) paste0("dissociation\nrate: ", x, " s-1")

plot_data %>% 
  ggplot(aes(x = log2(k_cap_bind), y = log2(psr), color = as.factor(uorf_init))) +
  facet_wrap(~ k_abort, ncol = 2, labeller = labeller(k_abort = k_abort_labeller)) +
  geom_point(size = 1) +
  geom_line() +
  viridis::scale_color_viridis(discrete = T, end = 0.9) +
  scale_y_continuous(breaks = scales::pretty_breaks(n=3)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n=3)) +
  theme(axis.text.y = element_text(size = 7), axis.text.x = element_text(size = 7)) +
  labs(x = "ribosome loading (log2, s-1)", y = "main ORF protein\noutput (log2, s-1)",
       color = "Baseline uORF\ninitiation (%)")

ggsave('figures/collision_mediated_40S_dissociation_buffering.pdf', limitsize = FALSE)

```

# Get the rxns for scanning and initiating ribosomes at uORF2 start codons either 25 nt

```{r}
subset_rxns_uorf <- rxn_data %>% 
  filter(str_detect(name, "^(scan_[:digit:]|scan_from_|scan_to)")) %>% 
  filter(str_detect(name, c("_25$"))) %>%
  select(sim_id, name) %>%
  print()
```

# Get the number of scanning and initiating ribosomes at uORF2 start codon 25 nt and join with parameters

```{r}
uorf2_init_data <- rxn_data %>% 
  right_join(subset_rxns_uorf, by = c("sim_id", "name")) %>%
  select(sim_id, n_firings, name) %>% 
  spread(name, n_firings) %>% 
  left_join(param_data, by = c("sim_id")) %>%
  print()
```

# Get the fraction of scanning and initiating ribosomes at 25 nt uORF2 start codon

```{r}
uorf2_n_data_25 <- uorf2_init_data %>%
  mutate(total = scan_25 + scan_to_elongate_25 + scan_to_elongate_from_leading_collision_25) %>% 
  mutate(init_fraction = (scan_to_elongate_25 + scan_to_elongate_from_leading_collision_25) / total * 100) %>%
  print()
```

# Queuing-mediated enhanced repression uORF initiation faceted by d_stall

```{r,fig.width=3,fig.height=1.25}
plot_data <- data %>%
  left_join(uorf2_n_data_25, by = c("sim_id", "uorf_init", "k_elong_stall", "k_cap_bind", "l_scan_capture", "d_stall", "uorf_reinit")) %>%
  filter(k_elong_stall == 0.001,
         uorf_init != 0,
         uorf_reinit == 50,
         l_scan_capture == 0,
         k_abort == 0) %>%
  print()

d_stall_labeller <- function(x) paste0("d_stall: ", x, " nt")

plot_data %>% 
  ggplot(aes(x = log2(k_cap_bind), y = init_fraction, color = as.factor(uorf_init))) +
  facet_wrap(~ d_stall, ncol = 2, labeller = labeller(d_stall = d_stall_labeller)) +
  geom_point(size = 1) +
  geom_line() +
  viridis::scale_color_viridis(discrete = T, end = 0.9) +
  scale_y_continuous(breaks = scales::pretty_breaks(n=3)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n=3)) +
  theme(axis.text.y = element_text(size = 7), axis.text.x = element_text(size = 7)) +
  labs(x = "ribosome loading (log2, s-1)", y = "uORF\ninitiation (%)",
       color = "Baseline uORF\ninitiation (%)")

ggsave('figures/queuing_mediated_enhanced_repression_increased_init_from_queue.pdf')

```

# Plot PSR as a function of initiation rate for different rates of uORF2 initiation, multiple k_abort for collision-mediated 40S dissociation model

```{r,fig.width=2,fig.height=1.25}
plot_data <- data %>%
  filter(k_elong_stall == 0.001,
         uorf_init == 10,
         d_stall == 63,
         uorf_reinit == 50,
         k_abort != 0,
         l_scan_capture == 0) %>%
  print()

plot_data %>% 
  ggplot(aes(x = log2(k_cap_bind), y = log2(psr), color = as.factor(log2(k_abort)))) +
  geom_point(size = 1) +
  geom_line() +
  viridis::scale_color_viridis(discrete = T, end = 0.9) +
  scale_y_continuous(breaks = scales::pretty_breaks(n=3)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n=3)) +
  theme(axis.text.y = element_text(size = 7), axis.text.x = element_text(size = 7)) +
  labs(x = "ribosome loading\n(log2, s-1)", y = "main ORF protein\noutput (log2, s-1)",
       color = "dissociation\nrate (log2, s-1)")

ggsave('figures/collision_mediated_40S_dissociation_buffering_more_abort_rates.pdf', limitsize = FALSE)

```