---
title: "Analysis of computational simulations of backwards scanning in the queueing-mediated enhanced repression model"
author: "`r Sys.info()[['user']]`"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  github_document:
    toc: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

# Load libraries

```{r}
library(tidyverse)
library(rasilabRtemplates)
library(viridis)
library(dplyr)
# disable scientific notation
options(scipen=999)
```

# Read reaction firing data

```{r}
rxn_data <- list.files("output/", pattern = "rxn_list.tsv.gz", full.names = T) %>% 
  enframe() %>% 
  mutate(sim_id = as.integer(str_extract(value, "(?<=model_)\\d+"))) %>% 
  mutate(data = map(value, read_tsv)) %>%
  select(-name, -value) %>% 
  unnest() %>%
  print()
```

# Read simulation time data

```{r}
sim_data <- list.files("output/", pattern = "molecule_type_list.tsv.gz", full.names = T) %>% 
  enframe() %>% 
  mutate(sim_id = as.integer(str_extract(value, "(?<=model_)\\d+"))) %>% 
  mutate(data = map(value, read_tsv)) %>%
  select(-name, -value) %>% 
  unnest() %>% 
  filter(mol_type %in% c("simulated_time", "cpu_time")) %>% 
  spread(mol_type, mol_type_id) %>% 
  print()
```

# Read last reaction time data

```{r}
last_rxn_data <- list.files("output/", pattern = "molecule_type_list.tsv.gz", full.names = T) %>% 
  enframe() %>% 
  mutate(sim_id = as.integer(str_extract(value, "(?<=model_)\\d+"))) %>% 
  mutate(data = map(value, read_tsv)) %>%
  select(-name, -value) %>% 
  unnest() %>% 
  filter(mol_type == "last_rxn_firing_time") %>%
  spread(mol_type, mol_type_id) %>% 
  print()
```

# Check to make sure last reaction fired near end of simulation. 
0 rows output means all simulations' last reactions fired in last 0.1% of simulated time

```{r}
check_rxn <- last_rxn_data %>%
  left_join(sim_data, by = "sim_id") %>%
  mutate(ratio = last_rxn_firing_time / simulated_time) %>%
  filter(ratio < 0.999) %>%
  print()
```

# Read simulation parameters

```{r}
annotations  <- read_tsv('sim.params.tsv') %>%
  print()
```

# Calculate protein synthesis rate of main ORF

```{r}
psr_data <- rxn_data %>% 
  mutate(mrna_loc = str_extract(name, "\\d+$")) %>%
  left_join(annotations %>% select(sim_id, uorf2_stop, orf_stop), by = "sim_id") %>% 
  filter((mrna_loc == orf_stop) & str_detect(name, "terminate")) %>% 
  group_by(sim_id, mrna_loc) %>% 
  dplyr::summarize(n_firings = sum(n_firings)) %>%
  left_join(sim_data, by = "sim_id") %>% 
  mutate(psr = n_firings / simulated_time) %>% 
  select(sim_id, mrna_loc, psr) %>%
  print()
```

# Get only annotations that we want to use for plotting

```{r}
param_data <- annotations %>%
  mutate(uorf_init = round(100*k_start_uorf2 / (5 + k_start_uorf2 + k_backwards_scan)),2) %>%
  mutate(d_stall = x_stall - uorf2_start) %>%
  mutate(uorf_reinit = 100*5/(k_terminated_ssu_recycle_uorf2 + 5)) %>%
  select(sim_id, k_cap_bind, k_elong_stall, uorf_init, k_scan_term_3_hit_80s, d_stall, uorf_reinit, k_backwards_scan) %>%
  print()
```

# Combine all processed data into a single table

```{r}
data <- param_data %>% 
  left_join(psr_data, by = "sim_id") %>%
  dplyr::rename(k_abort = k_scan_term_3_hit_80s) %>%
  select(-k_elong_stall, -uorf_init, -uorf_reinit, -mrna_loc) %>%
  filter(k_abort == 0) %>%
  print()
```

# Plot PSR as a function of initiation rate for different rates of backwards scanning & dstall

```{r,fig.width=3.5,fig.height=1.25}
plot_data <- data %>%
  filter(k_backwards_scan %in% c(0, 0.625, 2.5, 3.75))

d_stall_labeller <- function(x) paste0("d_stall: ", x, " nt")

plot_data %>% 
  ggplot(aes(x = log2(k_cap_bind), y = log2(psr), color = as.factor(k_backwards_scan))) +
  facet_wrap(~ d_stall, ncol = 3, labeller = labeller(d_stall = d_stall_labeller)) +
  geom_point(size = 1) +
  geom_line() +
  viridis::scale_color_viridis(discrete = T, end = 0.9) +
  scale_y_continuous(breaks = scales::pretty_breaks(n=2)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n=3)) +
  theme(axis.text.y = element_text(size = 7), axis.text.x = element_text(size = 7)) +
  labs(x = "ribosome loading (log2, s-1)", y = "main ORF protein\noutput (log2, s-1)",
       color = "backwards\nscanning\nrate (nt/s)")

ggsave('figures/fib_s2d.pdf', limitsize = FALSE)

```
